---
title: "Breeding Game, Final Results"
author: "PlantBreedGame"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
params:
  progressBar: NULL
  evalT1: NA
  evalT2: NA
  evalT3: NA
  evalT1T2: NA
  breederHistoryTimeLines: NA
  genealogy: NA
  BVdta: NA
  selInds : NA
colorlinks: true
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: TRUE
    code_folding: hide
    df_print: paged

---

```{r setup, include=FALSE}
progressBar <- params$progressBar

if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Initialisation..."
  )
}

knitr::opts_chunk$set(
  echo = FALSE, # show code
  eval = TRUE, # eval code
  include = TRUE, # if false, show nothing
  warning = FALSE,
  message = FALSE,
  error = TRUE, # do not stop on error
  results = "hold", # display chunk output in one block

  # df_print paged options
  rows.print = 10,
  max.print = Inf,
  tidy = FALSE, # better code organisation
  tidy.opts = list(width.cutoff = 80),
  strip.white = TRUE, # remove the white lines in the beginning or end of a source chunk

  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)

knitr::opts_knit$set(
  progress = TRUE,
  verbose = FALSE,
  width = 80
)
```

```{css echo=FALSE, eval = TRUE}
/* add some CSS styling here */
```


```{r}
suppressPackageStartupMessages({
  library(plotly)
  library(htmltools)
})

#  R options
options(stringsAsFactors = FALSE)
```

Thank you for playing the PlanBreedGame ! There is a small report of the game session.

For more information about this game, please check [the web site](https://sourcesup.renater.fr/plantbreedgame/) and the [code repository](https://github.com/timflutre/PlantBreedGame).

#### Citation {-}

Flutre, T., Diot, J., and David, J. (2019). PlantBreedGame: A Serious Game that Puts Students in the Breeder’s Seat. Crop Science. DOI 10.2135/cropsci2019.03.0183le

#### Copyright {-}

2015-2019: [INRAE](www.inrae.fr) , [Montpellier SupAgro](https://www.montpellier-supagro.fr/) 



# Evaluation


## Trait 1, flower production in kg/ha 

```{r evalT1}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Evaluation Trait 1..."
  )
}

params$evalT1
```

## Trait 2, sepmetin content in g/kg

```{r evalT2}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Evaluation Trait 2..."
  )
}

params$evalT2
```

## Trait 3, presence of symptoms caused by P. psychedelica

In this plot, `+1` means presence of symptom and `-1` means abscence of symptoms.

```{r evalT3}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Evaluation Trait 3..."
  )
}

params$evalT3
```

## Trait 1 vs Trait 2

```{r evalT1T2}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Evaluation Trait 1 vs Trait 2..."
  )
}

params$evalT1T2
```

## Pedigree {.tabset .tabset-fade .tabset-pills}

```{r evalPed}
genealogy <- params$genealogy

if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Evaluation Pedigree..."
  )
}
```


```{r results='asis', fig.width = 15, fig.height = 15}
for (breeder in names(genealogy)) {
  if (breeder == "Initial collection") next()
  cat("\n### ", breeder, "\n")

  plotPedigree(genealogy[[breeder]]$child,
    genealogy[[breeder]]$parent1,
    genealogy[[breeder]]$parent2,
    genealogy[[breeder]]$generation,
    edge.col.mother = "black",
    edge.col.father = "black",
    vertex.label.color = "darkgreen",
    vertex.size = 15,
    main = breeder,
    plot.it = TRUE,
    verbose = 0
  )

  cat("\n")
}
```

## Request history {.tabset .tabset-fade .tabset-pills}

```{r}
breederHistoryTimeLines <- params$breederHistoryTimeLines

if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Request history..."
  )
}
```


```{r evalReqHist, results='asis'}
for (breeder in names(breederHistoryTimeLines)) {
  cat("### ", breeder, " {-}\n")
  cat(knitr::knit_print(breederHistoryTimeLines[[breeder]]))
  cat("\n\n")
}
```



















# Game progress overview

```{r include=FALSE}
BVdta <- params$BVdta
selInds <- params$selInds
BVdta$selected <- 0
apply(selInds, 1, function(l) {
  BVdta$selected[BVdta$breeder == l[1] & BVdta$ind == l[2]] <<- 1
  invisible()
})
```

## BV of all individuals {.tabset .tabset-fade .tabset-pills}

This is an interactive plot. You can show/hide breeders by clicking on the legend.

Individuals selected for the final evaluation are represented by stars.


### Yeild (Trait 1)

Stars are the individuals selected for evaluation.

```{r}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "All BV T1..."
  )
}

p <- plot_ly(
  type = "scatter",
  mode = "markers"
)

dta <- BVdta[BVdta$selected == 0, ]

p <- p %>% add_markers(
  data = dta,
  color = dta$breeder,
  x = jitter(dta$gen),
  y = dta$trait1,
  size = dta$selected / 5,
  opacity = 0.95,
  hoverinfo = "text",
  text = paste0(
    dta$breeder,
    dta$ind,
    "\nparent1: ", dta$parent1,
    "\nparent2: ", dta$parent2,
    "\nBV trait1 = ", round(dta$trait1, 2),
    "\nBV trait2 = ", round(dta$trait2, 2),
    "\nBV trait1 x trait2 = ", round(dta$t1t2, 2)
  )
)


dta <- BVdta[BVdta$selected == 1, ]
p <- p %>%
  add_markers(
    inherit = FALSE,
    data = dta,
    color = dta$breeder,
    marker = list(
      symbol = "star",
      size = 9,
      line = list(
        color = "rgb(0, 0, 0)",
        width = 1
      )
    ),
    x = jitter(dta$gen),
    y = dta$trait1,
    size = dta$selected / 5,
    opacity = 0.95,
    hoverinfo = "text",
    text = paste0(
      dta$breeder,
      "\n<b>", dta$ind, "*", "</b>",
      "\nparent1: ", dta$parent1,
      "\nparent2: ", dta$parent2,
      "\nBV trait1 = ", round(dta$trait1, 2),
      "\nBV trait2 = ", round(dta$trait2, 2),
      "\nBV trait1 x trait2 = ", round(dta$t1t2, 2)
    )
  ) %>%
  layout(
    title = paste0("All individuals (Trait 1)"),
    xaxis = list(
      title = "Generation",
      dtick = 1
    ),
    yaxis = list(
      title = "Breedind values"
    )
  )


p
```

### Sepmetine concentration (Trait 2)

```{r}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "All BV T2..."
  )
}

p <- plot_ly(
  type = "scatter",
  mode = "markers"
)

dta <- BVdta[BVdta$selected == 0, ]

p <- p %>% add_markers(
  data = dta,
  color = dta$breeder,
  x = jitter(dta$gen),
  y = dta$trait2,
  size = dta$selected / 5,
  opacity = 0.95,
  hoverinfo = "text",
  text = paste0(
    dta$breeder,
    dta$ind,
    "\nparent1: ", dta$parent1,
    "\nparent2: ", dta$parent2,
    "\nBV trait1 = ", round(dta$trait1, 2),
    "\nBV trait2 = ", round(dta$trait2, 2),
    "\nBV trait1 x trait2 = ", round(dta$t1t2, 2)
  )
)


dta <- BVdta[BVdta$selected == 1, ]
p <- p %>%
  add_markers(
    inherit = FALSE,
    data = dta,
    color = dta$breeder,
    marker = list(
      symbol = "star",
      size = 9,
      line = list(
        color = "rgb(0, 0, 0)",
        width = 1
      )
    ),
    x = jitter(dta$gen),
    y = dta$trait2,
    size = dta$selected / 5,
    opacity = 0.95,
    hoverinfo = "text",
    text = paste0(
      dta$breeder,
      "\n<b>", dta$ind, "*", "</b>",
      "\nparent1: ", dta$parent1,
      "\nparent2: ", dta$parent2,
      "\nBV trait1 = ", round(dta$trait1, 2),
      "\nBV trait2 = ", round(dta$trait2, 2),
      "\nBV trait1 x trait2 = ", round(dta$t1t2, 2)
    )
  ) %>%
  layout(
    title = paste0("All individuals (Trait 1)"),
    xaxis = list(
      title = "Generation",
      dtick = 1
    ),
    yaxis = list(
      title = "Breedind values"
    )
  )
p
```

## Box plot {.tabset .tabset-fade .tabset-pills}

### Trait 1

```{r}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "All BV box T1..."
  )
}

plot_ly(
  data = BVdta,
  type = "box",
  x = ~gen,
  y = ~trait1,
  hoverinfo = "y",
  color = ~breeder
) %>%
  layout(
    boxmode = "group",
    title = paste0("All individuals (Trait1)"),
    xaxis = list(
      title = "Generation",
      dtick = 1
    ),
    yaxis = list(
      title = "Breedind values"
    )
  )
```



### Trait 2

```{r}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "All BV box T2..."
  )
}
plot_ly(
  data = BVdta,
  type = "box",
  x = ~gen,
  y = ~trait2,
  hoverinfo = "y",
  color = ~breeder
) %>%
  layout(
    boxmode = "group",
    title = paste0("All individuals (Trait2)"),
    xaxis = list(
      title = "Generation",
      dtick = 1
    ),
    yaxis = list(
      title = "Breedind values"
    )
  )
```



## Best individual for each trait {.tabset .tabset-fade .tabset-pills}

### Trait 1

```{r}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Best individuals T1..."
  )
}


# get the best individuals per breeder per generation
bestBV <- NULL
for (breeder in unique(BVdta$breeder)) {
  tmp1 <- BVdta[BVdta$breeder == breeder, ]
  for (gen in unique(tmp1$gen)) {
    tmp2 <- tmp1[tmp1$gen == gen, ]
    bestBV <- rbind(
      bestBV,
      tmp2[tmp2$trait1 == max(tmp2$trait1), ]
    )
  }

  # add initial collection ind to all breeders
  if (breeder != "Initial collection") {
    bestIni <- bestBV[bestBV$breeder == "Initial collection", ]
    bestIni$breeder <- breeder
    bestBV <- rbind(bestBV, bestIni)
  }
}

# remove "Initial collection" breeder
bestBV <- bestBV[bestBV$breeder != "Initial collection", ]


plot_ly(
  data = bestBV[order(bestBV$gen), ],
  type = "scatter",
  mode = "lines+markers",
  x = ~gen,
  y = ~trait1,
  color = ~breeder,
  hoverinfo = "text",
  text = ~ paste0(
    "<b>", ind, "</b>", # (in bold)
    "\nparent1: ", parent1,
    "\nparent2: ", parent2,
    "\nBV trait1 = ", round(trait1, 2),
    "\nBV trait2 = ", round(trait2, 2),
    "\nBV trait1 x trait2 = ", round(t1t2, 2)
  )
) %>%
  layout(
    title = paste0("Best individuals (Trait 1)"),
    xaxis = list(
      title = "Generation",
      dtick = 1
    ),
    yaxis = list(
      title = "Breedind values"
    )
  )
```

### Trait 2

```{r}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Best individuals T2..."
  )
}

# get the best individuals per breeder per generation
bestBV <- NULL
for (breeder in unique(BVdta$breeder)) {
  tmp1 <- BVdta[BVdta$breeder == breeder, ]
  for (gen in unique(tmp1$gen)) {
    tmp2 <- tmp1[tmp1$gen == gen, ]
    bestBV <- rbind(
      bestBV,
      tmp2[tmp2$trait2 == max(tmp2$trait2), ]
    )
  }

  # add initial collection ind to all breeders
  if (breeder != "Initial collection") {
    bestIni <- bestBV[bestBV$breeder == "Initial collection", ]
    bestIni$breeder <- breeder
    bestBV <- rbind(bestBV, bestIni)
  }
}

# remove "Initial collection" breeder
bestBV <- bestBV[bestBV$breeder != "Initial collection", ]


plot_ly(
  data = bestBV[order(bestBV$gen), ],
  type = "scatter",
  mode = "lines+markers",
  x = ~gen,
  y = ~trait2,
  color = ~breeder,
  hoverinfo = "text",
  text = ~ paste0(
    "<b>", ind, "</b>", # (in bold)
    "\nparent1: ", parent1,
    "\nparent2: ", parent2,
    "\nBV trait1 = ", round(trait1, 2),
    "\nBV trait2 = ", round(trait2, 2),
    "\nBV trait1 x trait2 = ", round(t1t2, 2)
  )
) %>%
  layout(
    title = paste0("Best individuals (Trait 2)"),
    xaxis = list(
      title = "Generation",
      dtick = 1
    ),
    yaxis = list(
      title = "Breedind values"
    )
  )
```



### Product Trait 2 x Trait 1


```{r}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Best individuals T1xT2..."
  )
}

# get the best individuals per breeder per generation
bestBV <- NULL
for (breeder in unique(BVdta$breeder)) {
  tmp1 <- BVdta[BVdta$breeder == breeder, ]

  for (gen in unique(tmp1$gen)) {
    tmp2 <- tmp1[tmp1$gen == gen, ]
    bestBV <- rbind(
      bestBV,
      tmp2[tmp2$t1t2 == max(tmp2$t1t2), ]
    )
  }

  # add initial collection ind to all breeders
  if (breeder != "Initial collection") {
    bestIni <- bestBV[bestBV$breeder == "Initial collection", ]
    bestIni$breeder <- breeder
    bestBV <- rbind(bestBV, bestIni)
  }
}

# remove "Initial collection" breeder
bestBV <- bestBV[bestBV$breeder != "Initial collection", ]


plot_ly(
  data = bestBV[order(bestBV$gen), ],
  type = "scatter",
  mode = "lines+markers",
  x = ~gen,
  y = ~t1t2,
  color = ~breeder,
  hoverinfo = "text",
  text = ~ paste0(
    "<b>", ind, "</b>", # (in bold)
    "\nparent1: ", parent1,
    "\nparent2: ", parent2,
    "\nBV trait1 = ", round(trait1, 2),
    "\nBV trait2 = ", round(trait2, 2),
    "\nBV trait1 x trait2 = ", round(t1t2, 2)
  )
) %>%
  layout(
    title = paste0("Best individuals (Trait 2)"),
    xaxis = list(
      title = "Generation",
      dtick = 1
    ),
    yaxis = list(
      title = "Breedind values"
    )
  )
```


## T1 vs T2 {.tabset .tabset-fade .tabset-pills}

```{r}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "T1vsT2..."
  )
}

myColors <- c(
  "#bbbbbb", "#165b16", "#0c246f", "#da4a3b", "#ffbe00",
  "#03A13D", "#7E28AD", "#1ED5E6", "#C78A1A", "#BF1B26"
)



# p <- htmltools::tagList()
p <- list()
for (breeder in unique(BVdta$breeder)) {
  # extract BV of the requested breeder
  if (breeder == "Initial collection") next()
  dta <- BVdta[BVdta$breeder %in% c(breeder, "Initial collection"), ]
  dta$gen <- sprintf("%02d", dta$gen)


  plt <- plot_ly(
    data = dta,
    type = "scatter",
    mode = "markers",
    colors = myColors
  )
  dta2 <- dta[dta$selected == 0, ]

  plt <- plt %>% add_markers(
    data = dta2,
    x = ~trait1,
    y = ~trait2,
    color = ~gen,
    size = ~ selected / 5,
    opacity = 0.95,
    hoverinfo = "text",
    text = ~ paste0(
      ind,
      "\nparent1: ", parent1,
      "\nparent2: ", parent2,
      "\nBV trait1 = ", round(trait1, 2),
      "\nBV trait2 = ", round(trait2, 2),
      "\nBV trait1 x trait2 = ", round(t1t2, 2)
    )
  )

  dta2 <- dta[dta$selected == 1, ]

  plt <- plt %>%
    add_markers(
      data = dta2,
      x = ~trait1,
      y = ~trait2,
      color = ~gen,
      marker = list(
        symbol = "star",
        size = 9,
        line = list(
          color = "rgb(0, 0, 0)",
          width = 1
        )
      ),
      size = ~ selected / 5,
      opacity = 0.95,
      hoverinfo = "text",
      text = ~ paste0(
        "\n<b>", ind, ifelse(selected == 1, "*", ""), "</b>", # (in bold)
        "\nparent1: ", parent1,
        "\nparent2: ", parent2,
        "\nBV trait1 = ", round(trait1, 2),
        "\nBV trait2 = ", round(trait2, 2),
        "\nBV trait1 x trait2 = ", round(t1t2, 2)
      )
    ) %>%
    layout(
      title = paste0("Trait 1 vs Trait 2 (", breeder, ")"),
      xaxis = list(
        title = "Trait 1"
      ),
      yaxis = list(
        title = "Trait 2"
      )
    )


  p[[breeder]] <- plt
}
```

```{r, results='asis'}
for (breeder in unique(BVdta$breeder)) {
  if (breeder == "Initial collection") next()
  cat("### ", breeder, "\n")
  # print(summary(lm(data=mtcars, mtcars[,1]~ mtcars[,i])))
  # print(p[[breeder]])
  cat(knitr::knit_print(p[[breeder]]))
  # print(as_widget(p[[breeder]]))
  cat("\n\n")
}
```


```{r}
if (!is.null(progressBar)) {
  progressBar$set(
    value = progressBar$getValue() + 1,
    message = "Compile Rmd:",
    detail = "Finalization..."
  )
}
```
