---
title: "PlantBreedGame: setup"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
params:
  progressBar: NULL
  rng_seed: 1993
  cost.pheno.field: 50
  cost.pheno.patho: 0.1
  cost.allof: 0.1
  cost.autof: 0.25
  cost.haplodiplo: 1
  cost.geno.hd: 1
  cost.geno.ld: 0.5
  cost.geno.single: 0.02
  cost.register: 4
  initialBudget: 3900
  t1_mu: 100
  t1_min: 20
  t1_cv_g: 0.1
  t1_h2: 0.3
  t2_mu: 15
  t2_min: 5
  t2_cv_g: 0.06
  t2_h2: 0.6
  prop_pleio: 0.4
  cor_pleio: -0.7
  max.nb.haplodiplos: 300
  max.nb.pltmatReq: 300
  nb.plots: 300
  maxEvalInds: 5
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: TRUE
    code_folding: hide
    self_contained: yes
  pdf_document:
    toc: true
    toc_depth: 4
    number_sections: TRUE
---

```{r setup, include=FALSE}
progressBar <- params$progressBar

if (is.null(progressBar)) {
  progressBar <- list(
    set = function(...){return(invisible(NULL))},
    getValue = function(...){return(invisible(NULL))}
  )
}

R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if (R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15)) {
  stop("requires R >= 2.15", call. = FALSE)
}

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  cache = FALSE,
  fig.align = "center",
  results = "hold",
  strip.white = TRUE
)

options(stringsAsFactors = TRUE) # R-4.0.0 compatibility

t0 <- proc.time()
```


# Overview

This document was conceived with the help of Jacques David (Montpellier SupAgro) and Philippe Brabant (AgroParisTech) for a master-level [workshop on genomic prediction and selection](https://github.com/timflutre/atelier-prediction-genomique).
It sets up a [serious game](https://sourcesup.renater.fr/plantbreedgame/) to teach plant selective breeding.

* Copyright 2016-2019 INRA

* License: [CC BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/)


# Set up the environment

```{r load_pkgs, warning = FALSE}
progress_detail = "Load R packages..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

suppressPackageStartupMessages(library(parallel))
nb.cores <- 1
if (Sys.info()["sysname"] != "Windows") {
  nb.cores <- 1

  ## NOTE: To be faster one can uncomment the code below but the code will be reproductible only if the same number of cores is used.
  # freeMem <- as.numeric(system("awk '/MemFree/ {print $2}' /proc/meminfo", intern = T))
  # maxCore <- max(floor(freeMem / (5 * 10^6)), 1)
  # availCore <- max(detectCores() - 1, 1)
  # nb.cores <- min(availCore, maxCore)
}
suppressPackageStartupMessages(library(DBI)) # from CRAN
suppressPackageStartupMessages(library(RSQLite)) # from CRAN
stopifnot(compareVersion(
  "2.0",
  as.character(packageVersion("RSQLite"))
)
!= 1)
suppressPackageStartupMessages(library(digest)) # from CRAN
suppressPackageStartupMessages(library(MASS)) # from CRAN
suppressPackageStartupMessages(library(rutilstimflutre)) # from GitHub
stopifnot(compareVersion(
  "0.170.0",
  as.character(packageVersion("rutilstimflutre"))
)
!= 1)
suppressPackageStartupMessages(library(scrm)) # from CRAN
suppressPackageStartupMessages(library(GenomicRanges)) # from bioconductor
source("src/fun/functions.R")
source("src/fun/module_gameInit_params.R")
source("src/fun/func_gameInit_validation.R")
source("src/fun/func_dbRequests.R")

session_id <- uuid::UUIDgenerate()
```

This new session have id: ` `r session_id` `

# Parameters

Validation and overview of the parameters used to initialised this game session.

```{r}
progress_detail = "Parameter validation..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

invisible({
  valid_positive_integer(params$rng_seed, strict = FALSE, accept_null = FALSE, raise_error = TRUE)

  valid_positive_number(params$cost.pheno.field, accept_null = FALSE, raise_error = TRUE)
  valid_positive_number(params$cost.pheno.patho, accept_null = FALSE, raise_error = TRUE)
  valid_positive_number(params$cost.allof, accept_null = FALSE, raise_error = TRUE)
  valid_positive_number(params$cost.autof, accept_null = FALSE, raise_error = TRUE)
  valid_positive_number(params$cost.haplodiplo, accept_null = FALSE, raise_error = TRUE)
  valid_positive_number(params$cost.geno.hd, accept_null = FALSE, raise_error = TRUE)
  valid_positive_number(params$cost.geno.ld, accept_null = FALSE, raise_error = TRUE)
  valid_positive_number(params$cost.geno.single, accept_null = FALSE, raise_error = TRUE)
  valid_positive_number(params$cost.register, accept_null = FALSE, raise_error = TRUE)
  valid_positive_number(params$initialBudget, accept_null = FALSE, raise_error = TRUE)

  valid_mu(params$t1_mu, accept_null = FALSE, raise_error = TRUE)
  valid_Tmin(params$t1_min, params$t1_mu, accept_null = FALSE, raise_error = TRUE)
  valid_cv_g(params$t1_cv_g, accept_null = FALSE, raise_error = TRUE)
  valid_h2(params$t1_h2, accept_null = FALSE, raise_error = TRUE)

  valid_mu(params$t2_mu, accept_null = FALSE, raise_error = TRUE)
  valid_Tmin(params$t2_min, params$t2_mu, accept_null = FALSE, raise_error = TRUE)
  valid_cv_g(params$t2_cv_g, accept_null = FALSE, raise_error = TRUE)
  valid_h2(params$t2_h2, accept_null = FALSE, raise_error = TRUE)

  t1_sig_p2 <- calc_sigma_p2(params$t1_mu, params$t1_min)
  valid_variance(t1_sig_p2, "phenotypes T1", accept_na = FALSE, accept_null = FALSE, raise_error = TRUE)
  t1_sig_a2 <- calc_sigma_a2(params$t1_cv_g, params$t1_mu)
  valid_variance(t1_sig_a2, "genetic values T1", accept_na = FALSE, accept_null = FALSE, raise_error = TRUE)
  t1_sig2 <- calc_sigma2(params$t1_h2, t1_sig_a2)
  valid_variance(t1_sig2, "noise T1", accept_na = FALSE, accept_null = FALSE, raise_error = TRUE)
  t1_sig_y2 <- calc_sigma_y2(t1_sig_p2, t1_sig_a2, t1_sig2 )
  valid_variance(t1_sig_y2, "year effects T1", accept_na = FALSE, accept_null = FALSE, raise_error = TRUE)

  t2_sig_p2 <- calc_sigma_p2(params$t2_mu, params$t2_min)
  valid_variance(t2_sig_p2, "phenotypes T2", accept_na = FALSE, accept_null = FALSE, raise_error = TRUE)
  t2_sig_a2 <- calc_sigma_a2(params$t2_cv_g, params$t2_mu)
  valid_variance(t2_sig_a2, "genetic values T2", accept_na = FALSE, accept_null = FALSE, raise_error = TRUE)
  t2_sig2 <- calc_sigma2(params$t2_h2, t2_sig_a2)
  valid_variance(t2_sig2, "noise T2", accept_na = FALSE, accept_null = FALSE, raise_error = TRUE)
  t2_sig_y2 <- calc_sigma_y2(t2_sig_p2, t2_sig_a2, t2_sig2 )
  valid_variance(t2_sig_y2, "year effects T2", accept_na = FALSE, accept_null = FALSE, raise_error = TRUE)

  valid_prop_pleio(params$prop_pleio, accept_null = FALSE, raise_error = TRUE)
  valid_cor_pleio(params$cor_pleio, accept_null = FALSE, raise_error = TRUE)

  valid_positive_integer(params$max.nb.haplodiplos, strict = TRUE, accept_null = FALSE, raise_error = TRUE)
  valid_positive_integer(params$max.nb.pltmatReq, strict = TRUE, accept_null = FALSE, raise_error = TRUE)
  valid_positive_integer(params$nb.plots, strict = TRUE, accept_null = FALSE, raise_error = TRUE)
  valid_positive_integer(params$maxEvalInds, strict = TRUE, accept_null = FALSE, raise_error = TRUE)
})

print(params[names(params) != "progressBar"])
```

# Set up data directories and RNG seed

```{r setup_dir}
progress_detail = "Setup directories..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)
data_root <- normalizePath(Sys.getenv("PLANTBREEDGAME_DATA_ROOT"))

root.dir <- file.path(data_root, session_id)
truth.dir <- file.path(root.dir, "truth")
shared.dir <- file.path(root.dir, "shared")
init.dir <- file.path(shared.dir, "initial_data")
reports.dir <- file.path(root.dir, "reports")
dbname <- file.path(root.dir, "breeding-game.sqlite")

if (dir.exists(root.dir)) {
  if (length(list.files(root.dir)) != 0) {
    stop(paste0("`data` directory (", root.dir, ") is not empty. Please empty this directory before running this script."))
  }
} else {
  dir.create(root.dir)
}

dir.create(truth.dir)
dir.create(shared.dir)
dir.create(init.dir)
dir.create(reports.dir)
```

```{r}
RNGkind("L'Ecuyer-CMRG") # to make mclapply reproducible
set.seed(params$rng_seed)
```

# Create initial breeders

Create the "admin" and "test" players (other players are created via the interface):
```{r}
breeders <- c("admin", "test")
breeder.dirs <- c()
for (breeder in breeders) {
  truth.breeder.dir <- paste0(truth.dir, "/", breeder)
  if (!dir.exists(truth.breeder.dir)) {
    dir.create(truth.breeder.dir)
  }
  breeder.dirs[breeder] <- paste0(shared.dir, "/", breeder)
  if (!dir.exists(breeder.dirs[breeder])) {
    dir.create(breeder.dirs[breeder])
  }
}
```

# Simulate haplotypes and genotypes

## Use the sequential coalescent with recombination

Nb of initial individuals:
```{r init_inds}
I <- 1000
ind.ids <- sprintf(fmt = paste0("ind%0", floor(log10(I)) + 1, "i"), 1:I)
```

Simulate haplotypes and genotypes of initial individuals:

- hypothesis: single pop

- hypothesis: no inter-chrom LD

- hypothesis: all chroms have the same length

- hypothesis: individuals are diploids

- scaling: measure time in units of `4 * N_e` generations


```{r init_chrs}
nb.chrs <- 10
L <- 10^6 # chromosome length, in base pairs
mu <- 10^(-8) # neutral mutation rate in events / base / generation
(u <- mu * L) # neutral mutation rate in events / chrom / gen
c <- mu # recomb rate in events / base / gen
(r <- c * L) # recomb rate in events / chrom / gen
```

Simplest evolutionary scenario similar to real data from domesticated plants:
```{r evol_scenario}
##                             _____
##                            /
##       ____________________/      Nec
##                       _Neb_
##                       |    \_____
##            Nea        |
##                       |
##                       |
##       ________________|
##                       2   1      0
## time  <---------------+---+------+
Nea <- 5 * 10^4 # ancestral
Neb <- 10 * 10^2 # bottleneck
Nec <- Ne0 <- 5 * 10^3 # current
(theta <- 4 * Ne0 * u) # scaled neutral mutation rate in events / chrom
(rho <- 4 * Ne0 * r) # scaled recomb rate in events / chrom
T1 <- 8000
T2 <- T1 + 2000
(alpha <- -(4 * Ne0) / T1 * log(Neb / Ne0)) # exp growth rate from Nec to Neb
(cmd <- paste0(
  "-eG ", T1 / (4 * Ne0), " ", alpha,
  " -eN ", T2 / (4 * Ne0), " ", Nea / Ne0
))
```

Enhancement: simulate two sub-populations, quite differenciated (by crossing individuals from each, the mean performance of their offsprings will be the same as if the parents were from the same pop, but their genetic variance will be larger)

Simulate genotypes via the coalescent:
```{r simul_geno_coalescent}

progress_detail = "Genotype simulation..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

f <- paste0(truth.dir, "/g0.RData")
g0 <- simulCoalescent(
  nb.inds = I,
  ind.ids = ind.ids,
  nb.reps = nb.chrs,
  pop.mut.rate = theta,
  pop.recomb.rate = rho,
  chrom.len = L,
  other = cmd,
  nb.pops = 1,
  get.trees = FALSE,
  get.tmrca = FALSE,
  permute.alleles = TRUE,
  verbose = 1
)
save(g0, file = f)
tools::md5sum(path.expand(f))
file.info(f)$mtime
head(g0$snp.coords)
g0$haplos$chr1[1:4, 1:6]
g0$genos[1:2, 1:6]
```

## Look at (M)AFs, LD, A

Look at the AFs and MAFs:
```{r look_mafs}
progress_detail = "AF calculation..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)
afs <- estimSnpAf(X = g0$genos, allow.updating = TRUE)
summary(afs)
plotHistAllelFreq(
  afs = afs,
  main = paste0(length(afs), " SNPs"),
  add.ml.beta = T
)
mafs <- estimSnpMaf(afs = afs)
summary(mafs)
sum(mafs >= 0.01) # nb of non-rare SNPs
plotHistMinAllelFreq(
  maf = mafs,
  main = paste0(length(mafs), " SNPs")
)
plotHistMinAllelFreq(
  maf = mafs[mafs >= 0.01],
  main = paste0(sum(mafs >= 0.01), " SNPs (MAF >= 0.01)")
)
```

Look at some haplotypes:
```{r look_haplos}
plotHaplosMatrix(g0$haplos$chr1[
  sort(sample.int(nrow(g0$haplos$chr1), 100)),
  mafs[1:ncol(g0$haplos$chr1)] >= 0.2
])
```

Estimate and plot pairwise LD per chromosome:
```{r estim_plot_ld}

progress_detail = "Generate LD plot..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

chr <- "chr1"
min.maf <- 0.2
min.pos <- 0
max.pos <- L
(length(snps.tokeep <- rownames(g0$snp.coords[mafs >= min.maf &
  g0$snp.coords$chr == chr &
  g0$snp.coords$pos >= min.pos &
  g0$snp.coords$pos <= max.pos, ])))
system.time(ld <- estimLd(
  X = g0$genos[, snps.tokeep],
  snp.coords = g0$snp.coords[snps.tokeep, ],
  use.ldcorsv = FALSE
))
nrow(ld)
summary(ld$cor2)
snp.dist <- distSnpPairs(
  snp.pairs = ld[, c("loc1", "loc2")],
  snp.coords = g0$snp.coords[snps.tokeep, ]
)
plotLd(snp.dist,
  sqrt(ld$cor2),
  estim = "r",
  main = paste0(
    length(snps.tokeep), " SNPs with MAF >= ", min.maf,
    " on ", chr
  ),
  use.density = TRUE,
  span = 1 / 20,
  sample.size = 2 * I,
  Ne = Ne0, c = c,
  add.ohta.kimura = TRUE
)
abline(v = 500, lty = 2)
```

Look at distances between consecutive SNPs:
```{r dist_btw_snps}

progress_detail = "SNP distances calculation..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

chr <- "chr3"
system.time(
  tmp <- distConsecutiveSnps(snp.coords = g0$snp.coords, only.chr = chr)
)
hist(tmp[[chr]],
  breaks = "FD", xlab = "in bp", las = 1,
  main = "Distances between consecutive SNPs"
)
abline(v = 500, lty = 2)
```

Thin SNPs:
```{r thin}
thin.snps <- thinSnps(
  method = "coord", threshold = 5 * 10^4,
  snp.coords = g0$snp.coords[names(mafs)[mafs >= 0.1], ]
)
length(thin.snps)
```

Estimate additive genomic relationships:
```{r estim_genrel}
A.vr <- estimGenRel(X = g0$genos[, thin.snps], method = "vanraden1")
A <- A.vr
imageWithScale(z = A, main = "Additive genomic relationships (A)")
inbred.coefs <- diag(A) - 1
summary(inbred.coefs) # compare to 0 if no inbreeding
hist(inbred.coefs, breaks = "FD")
abline(v = 0, lty = 2, col = "red")
coancestries <- A / 2
summary(diag(coancestries)) # compare to 0.5 for an outbred individual
hist(diag(coancestries), breaks = "FD", xlim = c(0, 1))
abline(v = 0.5, lty = 2, col = "red")
summary(coancestries[upper.tri(coancestries)])
hist(coancestries[upper.tri(coancestries)], breaks = "Sturges")
abline(v = 0, lty = 2, col = "red")
```

## Structure into families

Enhancement: panmixia with, say, 40 genotypes for 3 generations

## Choose SNPs for genotyping

Ascertain SNPs for the genotyping arrays (high and low density):

- hypothesis: mimick de novo sequencing of a few individuals

```{r ascertain_snps}
nb.inds.denovo <- 20
nb.snps.hd <- 10 * 10^3
nb.snps.ld <- floor(nb.snps.hd / 3)
subset.inds <- sort(sample(rownames(g0$genos), nb.inds.denovo))
mafs.subset <- estimSnpMaf(g0$genos[subset.inds, ])
summary(mafs.subset)
sum(mafs.subset >= 0.1)
subset.snps <- list()
subset.snps[["hd"]] <- sample(names(mafs.subset[mafs.subset >= 0.1]), nb.snps.hd)
subset.snps[["ld"]] <- sample(subset.snps[["hd"]], nb.snps.ld)
summary(mafs[subset.snps[["hd"]]])
summary(mafs[subset.snps[["ld"]]])
f <- paste0(init.dir, "/snp_coords_hd.txt.gz")
write.table(
  x = g0$snp.coords[subset.snps[["hd"]], ],
  file = gzfile(f), quote = FALSE,
  sep = "\t", row.names = TRUE, col.names = TRUE
)
f <- paste0(init.dir, "/snp_coords_ld.txt.gz")
write.table(
  x = g0$snp.coords[subset.snps[["ld"]], ],
  file = gzfile(f), quote = FALSE,
  sep = "\t", row.names = TRUE, col.names = TRUE
)
```

## Make lines

Make collection of lines via haplodiploidization:
```{r coll_lines}

progress_detail = "Generate initial collection's genotype..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

f <- paste0(truth.dir, "/coll.RData")
coll <- list()
crosses <- data.frame(
  parent1 = ind.ids,
  parent2 = NA,
  child = sprintf(
    fmt = paste0("Coll%0", floor(log10(I)) + 1, "i"),
    1:I
  ),
  stringsAsFactors = FALSE
)
loc.crossovers <- drawLocCrossovers(crosses, sapply(g0$haplos, ncol))
system.time(
  coll$haplos <- makeCrosses(
    haplos = g0$haplos, crosses = crosses,
    loc.crossovers = loc.crossovers,
    howto.start.haplo = 0, nb.cores = nb.cores
  )
)
coll$genos <- segSites2allDoses(
  seg.sites = coll$haplos, ind.ids = crosses$child,
  snp.ids = colnames(g0$genos)
)
coll$snp.coords <- g0$snp.coords
save(coll, file = f)
tools::md5sum(path.expand(f))
file.info(f)$mtime
```

Free some memory:
```{r free_mem_genos}
rm(
  g0, ld, snp.dist, mafs, loc.crossovers, mafs.subset, crosses,
  coancestries, A, A.vr
)
gc()
```


# Simulate phenotypes

## Generic preparation

Make data frame encoding the design for the initial phenotypes given to the players:
```{r init_pheno}

progress_detail = "Simulate initial phenotypes..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

nb.lines.per.year <- 150
nb.plots.per.line.per.year <- 2
line.ids <- getIndNamesFromHaplos(coll$haplos)[1:(150 + (10 - 1) * 75)]
nb.years <- 10
first.year <- as.numeric(format(Sys.Date(), "%Y"))
first_pheno_year <- first.year - nb.years
p0.df <- makeDfInitPhenos(
  nb.lines.per.year = nb.lines.per.year, nb.years = nb.years,
  nb.plots.per.line.per.year = nb.plots.per.line.per.year,
  first.year = first_pheno_year, line.ids
)
head(p0.df)
str(p0.df)
summary(as.vector(table(p0.df$ind)))
```

Enhancement: arrange plots on a grid and return their coordinates

Export genotypes for the players using high-density chip, but only for a subset of lines:
```{r export_geno}
initial_genotype_file <- paste0(init.dir, "/Result_genos_subset-initialColl-hd.txt.gz")
subset.years <- as.numeric(levels(p0.df$year)[
  (floor(0.5 * nlevels(p0.df$year))):nlevels(p0.df$year)
])
initially_genotyped_inds <- unique(as.character(p0.df$ind[p0.df$year %in% subset.years]))
write.table(
  x = coll$genos[initially_genotyped_inds, subset.snps[["hd"]]],
  file = gzfile(initial_genotype_file), quote = FALSE,
  sep = "\t", row.names = TRUE, col.names = TRUE
)
```

Thin to avoid LD between causal SNPs? No, keep all SNPs for the moment.
```{r choose_causal_snps}
## mafs <- estimSnpMaf(X=coll$genos[levels(p0.df$ind),])
## thin.snps <- thinSnps(method="coord", threshold=5*10^2,
##                       snp.coords=coll$snp.coords[names(mafs)[mafs >= 0.01],])
length(thin.snps <- colnames(coll$genos))
```

## Hypotheses

Simulate phenotypes of initial lines:

* only 3 traits are considered

* trait 1 has additive infinitesimal genetic architecture (e.g. yield)

* trait 2 has additive infinitesimal genetic architecture (e.g. quality)

* for traits 1 and 2, all SNPs (even the rarest) have a non-zero effect

* traits 1 and 2 are negatively genetically correlated (i.e. pleiotropy)

* "year" effects are different for traits 1 and 2, with possibly different variances, but no covariance between traits nor plots

* there is no "plot" effect, nor spatial heterogeneity

* errors are different for traits 1 and 2, with possibly different variances, but no covariance between traits, plots nor years

* pathogens present every three years

* trait 3 has a single, major QTL (e.g. resistance)

* "trait3=0" means "absence of symptoms" and, very likely, presence of the resistance allele (but the years with pathogen, some genotypes without the resistant allele may exceptionnally show no symptom)

* the causal SNP is on the low-density chip

* the resistance-confering allele is the minor one, but isn't too rare

* the resistance-confering allele has a dominant effect

## Notations

* $t \in \{1,\ldots,T\}$: index of the traits, with $T = 2$

* $i \in \{1,\ldots,I\}$: index of the genotypes, with $I$ = `r nlevels(p0.df$ind)`

* $j \in \{1,\ldots,J\}$: index of the years, with $J$ = `r nlevels(p0.df$year)`

* $k \in \{1,\ldots,K\}$: index of the plots, with $K$ = `r nlevels(p0.df$plot)`

* $n \in \{1,\ldots,N\}$: index of the phenotypes, with $N$ = `r nrow(p0.df)`

* $p \in \{1,\ldots,P\}$: index of the SNPs, with $P$ = `r length(thin.snps)`

* $y_{ijk}^{(t)}$: phenotype of genotype $i$, year $j$ in plot $k$ for trait $t$, with variance $\sigma_p^{2 \, (t)}$

* $\alpha_j^{(t)}$: effect of year $j$ on trait $t$, with variance $\sigma_\alpha^{2 \, (t)}$

* $g_i^{(t)}$: genotypic value of genotype $i$ for trait $t$, with variance $\sigma_g^{2 \, (t)}$, assuming only additive contributions, that is $g_i^{(t)} = g_{A,i}^{(t)}$ with variance $\sigma_a^{2 \, (t)}$

* $\epsilon_{ijk}^{(t)}$: error, with variance $\sigma^{2 \, (t)}$

* $X$: $I \times P$ matrix containing the SNP genotypes coded in $\{0,1,2\}$

* $\beta_p^{(t)}$: additive SNP effect of SNP $p$ on trait $t$, with variance $\sigma_\beta^{2 \, (t)}$

## Models

### Generative model of the phenotypes

The generative model is used to simulate the data.
Let's write it generically for a given trait, hence discarding the subscript $^{(t)}$:

\[
\forall i, j, k, \; y_{ijk} = \mu + \alpha_j + g_i + \epsilon_{ijk}
\]

with:

* $\alpha_j \sim \mathcal{N}(0, \sigma_\alpha^2)$;

* $g_i = g_{A,i} \sim \mathcal{N}(0, \sigma_a^2)$;

* $\epsilon_{ijk} \sim \mathcal{N}(0, \sigma^2)$;

so that:

* $y_{ijk} \sim \mathcal{N}(\mu_p, \sigma_p^2)$;

* $\mathbb{E}[y_{ijk}] = \mu_p = \mu$;

* $\mathbb{V}[y_{ijk}] = \sigma_p^2 = \sigma_a^2 + \sigma_\alpha^2 + \sigma^2$.

See below for the procedure aiming at choosing the variances so that several constraints are satisfied.

### Statistical model of the phenotypes

The statistical model is used to perform inference, i.e. estimate the parameters and predict the random variables.
The likelihood can be written as:

\[
\forall i, j, k, \; y_{ijk} = \mu + \alpha_j + g_i + \epsilon_{ijk}
\]

with:

* $\alpha_j$ modeled as fixed effects as there is no genotype-year interaction;

* $g_i = g_{A,i} \sim \mathcal{N}(0, \sigma_a^2)$;

* $\epsilon_{ijk} \sim \mathcal{N}(0, \sigma^2)$;

so that:

* $y_{ijk} \sim \mathcal{N}(\mu_p, \sigma_p^2)$;

* $\mathbb{E}[y_{ijk}] = \mu_p = \mu + \alpha_j$;

* $\mathbb{V}[y_{ijk}] = \sigma_p^2 = \sigma_a^2 + \sigma^2$.

Assuming the design is balanced with each genotype observed in 2 plots per year, for 2 years (which is almost the case), the broad-sense/narrow-sense heritabilities of the design (i.e. heritabilities of the phenotypic means, also called "replicabilities") are defined as:

* $H_{\text{design}}^2 = \frac{\sigma_g^2}{\sigma_p^{2 \star}}$ where $\sigma_p^{2 \star} = \sigma_g^2 + \frac{\sigma^2}{n^\star}$ with, here, $n^\star \approx 4$;

* $h_{\text{design}}^2 = \frac{\sigma_a^2}{\sigma_p^{2 \star}}$.

### Genotypic models

The above generative model is parametrized in terms of genotypic values, the $\{g_i\}$'s.
However, these values must themselves be simulated from SNP genotypes, the $X$ matrix, and SNP effects, the $\{\beta_p\}$'s (as $g_i = g_{A,i}$, only the SNP genotypes encoded additively will be considered).

Hopefully, there exists an equivalence between both parametrizations, as long as certain requirements are satisfied.
Following [Habier et al (2007)](http://dx.doi.org/10.1534/genetics.107.081190):

* model RR (ridge regression): also called "RR-BLUP"
    + $\boldsymbol{g} = \boldsymbol{1} \mu^\ast + X \boldsymbol{\beta} + \boldsymbol{\epsilon}^\ast$
    + $\boldsymbol{\beta} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\beta^2 \, \text{Id})$
    + $\boldsymbol{\epsilon}^\ast \sim \mathcal{N}(\boldsymbol{0}, \sigma^{2\ast} \, \text{Id})$
    + $\Rightarrow \mathbb{E}[\boldsymbol{g}] = \boldsymbol{1} \mu^\ast$
    + $\Rightarrow \mathbb{V}[\boldsymbol{g}] = \sigma_\beta^2 \, X X^T + \sigma^{2\ast} \, \text{Id}$

* model AM ("animal model"): also called "GBLUP"
    + $\boldsymbol{g} = \boldsymbol{1} \mu^\ast + \boldsymbol{g}_A + \boldsymbol{\epsilon}^\ast$
    + $\boldsymbol{g}_A \sim \mathcal{N}(\boldsymbol{0}, \sigma_a^2 \, A)$ where $A$ is the matrix of additive genetic relationships (from the pedigree)
    + $\boldsymbol{\epsilon}^\ast \sim \mathcal{N}(\boldsymbol{0}, \sigma^{2\ast} \, \text{Id})$
    + $\Rightarrow \mathbb{E}[\boldsymbol{g}] = \boldsymbol{1} \mu^\ast$
    + $\Rightarrow \mathbb{V}[\boldsymbol{g}] = \sigma_a^2 \, A + \sigma^{2\ast} \, \text{Id}$

It can be shown that:

\[
\mathbb{E}[X X^T] = A \; \left( 2 \sum_p f_p (1 - f_p) \right) + \boldsymbol{1} \boldsymbol{1}^T 4 \sum_p f_p^2
\]

where the $\{f_p\}$'s are the SNP allele frequencies.

That is, $\mathbb{E}[X X^T]$ is proportional to $A$ apart from a constant.
Thus, models RR and AM are equivalent if one replaces $A$ by $\frac{X X^T}{2 \sum_p f_p (1 - f_p)}$, assuming a large enough number of SNPs.
Thus, knowing the additive genotypic variance, $\sigma_a^2$, we can deduce the variance of the SNP effects: $\sigma_\beta^2 = \frac{\sigma_a^2}{2 \sum_p f_p (1 - f_p)}$.

The constant mentioned above is the same for all genotypes, so it will end up in the intercept of the statistical model, $\hat{\mu}^\ast$.

Note that, when $\boldsymbol{g}$ is simulated as $X \boldsymbol{\beta}$ with $X$ in $\{0,1,2\}$, its mean isn't $\boldsymbol{0}$.
For this to be the case, the $X$ matrix should be centered using the allele frequencies from the whole population.
More details in [Vitezica et al (2013)](http://dx.doi.org/10.1534/genetics.113.155176) and [Toro et al (2011)](http://www.gsejournal.org/content/43/1/27).
In the course of the game, it would be better to update the allele frequencies each time new plant material is created.

### Selection model

As designer of the game, we can set the difficulty of the registration requirements by tuning the required response to selection, $R = \mu_1 - \mu$, where $\mu_1$ is the phenotypic mean (global intercept) at the final generation (just before registration) and $\mu$ is the mean of the initial collection.

We can also write $R$ as $w \times \mu$ where $w$ correspond to the registration threshold compare to the mean of the initial collection, and we also have the following relations:

* $R = h^2 \; i \; \sigma_p$;

* $h2 = \frac{\sigma_a^2}{\sigma_p^2}$.

Thus: $w \, \mu = \frac{\sigma_a^2}{\sigma_p^2} \, i \, \sigma_p \Leftrightarrow \frac{w}{i} = \frac{\sigma_a^2}{\sigma_p \, \mu}$

Given the definition of the genotypic coef of variation, $\text{CV_g} = \frac{\sigma_g}{\mu}$, we get:

\[
\frac{w}{i} = \frac{\mu}{\sigma_p} \, \text{CV_g}^2
\]

Therefore, making things more or less difficult boils down to fine-tuning $i$, which will be reflected in $\text{CV_g}$ (for $\mu$ and $\sigma_p$ fixed).

## Simulation procedure for trait 1

Here is the procedure for trait 1 of setting the variances required by the generative model above to sample data.

We set the initial phenotypic mean, $\mu_p$, to a global intercept which will be held constant during the whole game: $\mu_p = \mu$.

```{r simul_proc_t1_mu}
(mu <- params$t1_mu)
```

We set the individual-plant phenotypic variance so that approximately 99% of all the phenotypes are above a minimal value: $\mu - 3 \sigma_p \ge y_{\text{min}}$.

```{r simul_proc_t1_sigma_p2}
y.min <- params$t1_min
(sigma.p2 <- ((mu - y.min) / 3)^2)
sqrt(sigma.p2)
```

We set the registration threshold to be realistic:

```{r}
register.min.trait1 <- 1.03 # compare to the phenotypic mean of the controls
(w <- register.min.trait1 - 1)
```

We set the selection intensity, but note that the controls will be chosen (see below) so that they are among the best from the initial collection, hence the chosen selection intensity shouldn't be too large:

```{r}
i <- 1
## i <- 2
## i <- 3
## i <- 4
```

We deduce the genotypic coef of variation:

```{r}
(CV.g <- sqrt((w / i) * (sqrt(sigma.p2) / mu)))
```

However, because the previous computations don't account for the negative correlation between traits, we still fix the CVg by hand at this point:
```{r}
CV.g <- params$t1_cv_g
```

We deduce the (additive) genotypic variance:

```{r simul_proc_t1_sigma_a2}
(sigma.a2 <- (CV.g * mu)^2)
sqrt(sigma.a2)
stopifnot(sigma.a2 < sigma.p2)
```

We deduce the error variance given that the intra-annual, individual-plant heritability is set at a particular value: $h^2 = \frac{\sigma_a^2}{\sigma_a^2 + \sigma^2}$.

```{r simul_proc_t1_sigma2}
h2 <- params$t1_h2
(sigma2 <- ((1 - h2) / h2) * sigma.a2)
sqrt(sigma2)
stopifnot(sigma2 + sigma.a2 < sigma.p2)
```

From all this, we deduce the year variance: $\sigma_\alpha^2 = \sigma_p^2 - \sigma_a^2 - \sigma^2$.

```{r simul_proc_t1_sigma_alpha2}
(sigma.alpha2 <- sigma.p2 - sigma.a2 - sigma2)
sqrt(sigma.alpha2)
```

We can also compute the heritability of the design (from the statistical model above), for various number of plots per genotype per year:

```{r simul_proc_t1_H2_design}
(H2.design <- sigma.a2 / (sigma.a2 + sigma2 / (2 * mean(as.vector(table(p0.df$ind)))))) # compare with estimate in infer.Rmd
sigma.a2 / (sigma.a2 + sigma2 / (2 * 2))
sigma.a2 / (sigma.a2 + sigma2 / (2 * 3))
sigma.a2 / (sigma.a2 + sigma2 / (2 * 4))
sigma.a2 / (sigma.a2 + sigma2 / (2 * 5))
sigma.a2 / (sigma.a2 + sigma2 / (2 * 20)) # used for the evaluation
```

Based on the SNP genotypes, we can deduce the variance of the SNP effects, $\sigma_\beta^2$.

```{r simul_proc_t1_sigma_beta2}
(sigma.beta2 <- sigma.a2 / (2 * sum(afs[thin.snps] * (1 - afs[thin.snps]))))
(constant <- 4 * sum(afs[thin.snps]^2))
```

Save all these values in a list:
```{r simul_proc_t1_save_values}
params.trait1 <- list(
  mu = mu, y.min = y.min, sigma.p2 = sigma.p2,
  i = i, CV.g = CV.g, sigma.a2 = sigma.a2,
  h2 = h2, sigma2 = sigma2, sigma.alpha2 = sigma.alpha2,
  H2.design = H2.design, sigma.beta2 = sigma.beta2
)
```

## Simulation procedure for trait 2

Here is the procedure for trait 2 of setting the variances required by the generative model above to sample data.

We fix the initial phenotypic mean, $\mu_p$, to a global intercept which will be held constant during the whole game: $\mu_p = \mu$.

```{r simul_proc_t2_mu}
(mu <- params$t2_mu)
```

We deduce the individual-plant phenotypic variance so that approximately 99% of all the phenotypes are above a minimal value: $\mu - 3 \sigma_p \ge y_{\text{min}}$.

```{r simul_proc_t2_sigma_p2}
y.min <- params$t2_min
(sigma.p2 <- ((mu - y.min) / 3)^2)
sqrt(sigma.p2)
```

We set the registration threshold (for this trait, there is no need to set a target selection intensity as $w=0$):

```{r}
register.min.trait2 <- mu
```

We deduce the (additive) genotypic variance given that the genotypic coef of variation is set at a particular value:

```{r simul_proc_t2_sigma_a2}
CV.g <- params$t2_cv_g
(sigma.a2 <- (CV.g * mu)^2)
sqrt(sigma.a2)
stopifnot(sigma.a2 < sigma.p2)
```

We deduce the error variance given that the intra-annual, individual-plant heritability is set at a particular value: $h^2 = \frac{\sigma_a^2}{\sigma_a^2 + \sigma^2}$.

```{r simul_proc_t2_sigma2}
h2 <- params$t2_h2
(sigma2 <- ((1 - h2) / h2) * sigma.a2)
sqrt(sigma2)
stopifnot(sigma2 + sigma.a2 < sigma.p2)
```

From all this, deduce the year variance: $\sigma_\alpha^2 = \sigma_p^2 - \sigma_a^2 - \sigma^2$.

```{r simul_proc_t2_sigma_alpha2}
(sigma.alpha2 <- sigma.p2 - sigma.a2 - sigma2)
sqrt(sigma.alpha2)
```

We can also compute the heritability of the design (from the statistical model above), for various number of plots per genotype per year:

```{r simul_proc_t2_H2_design}
(H2.design <- sigma.a2 / (sigma.a2 + sigma2 / (2 * mean(as.vector(table(p0.df$ind)))))) # compare with estimate in infer.Rmd
sigma.a2 / (sigma.a2 + sigma2 / (2 * 2))
sigma.a2 / (sigma.a2 + sigma2 / (2 * 3))
sigma.a2 / (sigma.a2 + sigma2 / (2 * 4))
sigma.a2 / (sigma.a2 + sigma2 / (2 * 5))
sigma.a2 / (sigma.a2 + sigma2 / (2 * 20)) # used for the evaluation
```

Based on the SNP genotypes, we can deduce the variance of the SNP effects, $\sigma_\beta^2$.

```{r simul_proc_t2_sigma_beta2}
(sigma.beta2 <- sigma.a2 / (2 * sum(afs[thin.snps] * (1 - afs[thin.snps]))))
(constant <- 4 * sum(afs[thin.snps]^2))
```

Save all these values in a list:
```{r simul_proc_t2_save_values}
params.trait2 <- list(
  mu = mu, y.min = y.min, sigma.p2 = sigma.p2,
  CV.g = CV.g, sigma.a2 = sigma.a2,
  h2 = h2, sigma2 = sigma2, sigma.alpha2 = sigma.alpha2,
  H2.design = H2.design, sigma.beta2 = sigma.beta2
)
```

## Simulate traits 1 and 2

Simulate the SNP effects for traits 1 and 2 jointly:
```{r simul_snp_effects_trait1_trait2}
(sigma.beta2 <- c(
  trait1 = params.trait1$sigma.beta2,
  trait2 = params.trait2$sigma.beta2
))
snp.effects12 <- simulSnpEffectsTraits12(
  snp.ids = thin.snps,
  sigma.beta2 = sigma.beta2,
  prop.pleio = params$prop_pleio, cor.pleio = params$cor_pleio
)
```

Make phenotypes for traits 1 and 2 jointly:
```{r make_phenos_trait1_trait2}
mu.traits12 <- c(trait1 = params.trait1$mu, trait2 = params.trait2$mu)
sigma.traits12 <- c(
  trait1 = params.trait1$sigma2,
  trait2 = params.trait2$sigma2
)
sigma.alpha2 <- c(
  trait1 = params.trait1$sigma.alpha2,
  trait2 = params.trait2$sigma.alpha2
)
p0 <- simulTraits12(
  dat = p0.df, mu = mu.traits12,
  sigma.alpha2 = sigma.alpha2,
  X = coll$genos[levels(p0.df$ind), thin.snps],
  afs = afs[thin.snps],
  Beta = snp.effects12$Beta,
  sigma2 = sigma.traits12
)
```

Do some checks:
```{r plots_phenos_trait1_trait2}
hist(p0$G.A[, 1], breaks = "FD", las = 1, col = "grey", border = "white")
hist(p0$G.A[, 2], breaks = "FD", las = 1, col = "grey", border = "white")
regplot(p0$G.A[, 1], p0$G.A[, 2], xlab = "p0$G.A[,1]", ylab = "p0$G.A[,2]", las = 1)
summary(p0$Y[, 1])
summary(p0$Y[, 2])
hist(p0$Y[, 1], breaks = "FD", las = 1, col = "grey", border = "white")
boxplot(p0$Y[, 1] ~ p0.df$year, las = 1, main = "p0$Y[,1]")
abline(h = mean(p0$Y[, 1]), lty = 2)
hist(p0$Y[, 2], breaks = "FD", las = 1, col = "grey", border = "white")
boxplot(p0$Y[, 2] ~ p0.df$year, las = 1, main = "p0$Y[,2]")
abline(h = mean(p0$Y[, 2]), lty = 2)
regplot(p0$Y[, 1], p0$Y[, 2], xlab = "p0$Y[,1]", ylab = "p0$Y[,2]", las = 1)
plot(p0$Y[, 1], p0$Y[, 2], xlab = "p0$Y[,1]", ylab = "p0$Y[,2]", las = 1, type = "n")
for (j in 1:nlevels(p0.df$year)) {
  year <- levels(p0.df$year)[j]
  idx <- p0.df$year == year
  points(p0$Y[idx, 1], p0$Y[idx, 2], col = j)
  abline(lm(p0$Y[idx, 2] ~ p0$Y[idx, 1]), col = j)
}
```

## Simulate trait 3

Make the "pathogen" column:
```{r patho_years}
p0.df$pathogen <- FALSE
(years.with.patho <- seq(from = first_pheno_year, by = 3, length.out = 4))
p0.df$pathogen[p0.df$year %in% years.with.patho] <- TRUE
```

Make trait 3:
```{r make_trait3}
p0$trait3 <- simulTrait3(
  dat = p0.df,
  X = coll$genos[levels(p0.df$ind), thin.snps],
  afs = afs[thin.snps],
  subset.snps = subset.snps[["ld"]],
  prob.resist.no.qtl = 0.02
)
table(p0$trait3$y)
```

## Finalize

Fill "phenotype" data.frame for players:
```{r fill_pheno_players}
p0.df$trait1.raw <- p0$Y[, 1]
p0.df$trait2 <- p0$Y[, 2]
p0.df$trait3 <- p0$trait3$y
```

Adjust trait 1 for yield loss in case of pathogen presence:
```{r adjust_trai1}
p0$prop.yield.loss <- 0.2
p0.df$trait1 <- p0.df$trait1.raw
idx <- p0.df$pathogen & (p0.df$trait3 == 1)
p0.df$trait1[idx] <- (1 - p0$prop.yield.loss) * p0.df$trait1[idx]
hist(p0.df$trait1, breaks = "FD", las = 1, col = "grey", border = "white")
boxplot(p0.df$trait1 ~ p0.df$year, las = 1, main = "p0.df$trait1")
abline(h = mean(p0.df$trait1), lty = 2)
regplot(p0.df$trait1, p0.df$trait2,
  xlab = "p0.df$trait1", ylab = "p0.df$trait2",
  las = 1
)
```

Enhancement: add missing data

Save data:
```{r save_phenos}

progress_detail = "Save initial phenotypes..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

f <- paste0(truth.dir, "/p0.RData")
save(p0, file = f)
tools::md5sum(path.expand(f))
file.info(f)$mtime
```

# Controls and registration

## Choose controls

At the end of the game, each breeder has to submit at least one new variety.
For them to be registered, they should be "better" than controls.
The controls are chosen so that they have:

* an "additive genotypic yield" among the best lines,

* an "additive genotypic quality" above the mean of all lines,

* the resistant allele for some of them (if possible).

Retrieve the breeding values of each line in the collection:
```{r rank_genos}
BVs <- data.frame(
  trait1 = p0$mu[1] + p0$G.A[, 1],
  trait2 = p0$mu[2] + p0$G.A[, 2],
  trait3 = !coll$genos[levels(p0.df$ind), p0$trait3$qtn.id] %in%
    p0$trait3$resist.genos,
  row.names = levels(p0.df$ind)
)
summary(BVs)
```

Identify potential controls:
```{r identify_potential_controls, eval=TRUE}
BVs <- BVs[order(BVs$trait1, decreasing = TRUE), ]
prop.best.lines <- 0.1 # to be decreased, e.g. to 10% or even 5%
length(idx1 <- 1:ceiling(prop.best.lines * nrow(BVs)))
summary(BVs[idx1, ])
length(idx2 <- which(BVs$trait2 >= mean(BVs$trait2)))
potential.controls <- BVs[intersect(idx1, idx2), ]
summary(potential.controls)
```

Select five controls among them:
```{r choose_controls, eval=TRUE}
nb.controls <- 5
nb.ctls.resist <- 2
stopifnot(sum(potential.controls$trait3 == 0) >= nb.ctls.resist)
nb.ctls.sensitiv <- nb.controls - nb.ctls.resist
stopifnot(sum(potential.controls$trait3 != 0) >= nb.ctls.sensitiv)
length(pot.ctls.resist <- rownames(potential.controls)[potential.controls$trait3 == 0])
ctls.resist <- sample(pot.ctls.resist, nb.ctls.resist)
length(pot.ctls.sensitiv <- rownames(potential.controls)[potential.controls$trait3 != 0])
ctls.sensitiv <- sample(pot.ctls.sensitiv, nb.ctls.sensitiv)
controls <- c(ctls.resist, ctls.sensitiv)
BVs[controls, ]
```

Plot how good they are compare to the rest of the collection:
```{r plot_controls_phenos}
plot(x = p0$mu[1] + p0$G.A[, 1], y = p0$mu[2] + p0$G.A[, 2], las = 1, main = "Controls")
abline(v = mean(p0$mu[1] + p0$G.A[, 1]), h = mean(p0$mu[2] + p0$G.A[, 2]), lty = 2)
points(
  x = p0$mu[1] + p0$G.A[ctls.resist, 1],
  y = p0$mu[2] + p0$G.A[ctls.resist, 2],
  pch = 19, col = "green"
)
points(
  x = p0$mu[1] + p0$G.A[ctls.sensitiv, 1],
  y = p0$mu[2] + p0$G.A[ctls.sensitiv, 2],
  pch = 17, col = "green"
)
legend("topright",
  legend = c("resistant", "sensitive"), pch = c(19, 17),
  col = "green", bty = "n"
)
plot(x = p0.df$trait1, y = p0.df$trait2, main = "Controls")
points(
  x = p0.df$trait1[p0.df$ind %in% ctls.resist],
  y = p0.df$trait2[p0.df$ind %in% ctls.resist],
  pch = 19, col = "green"
)
points(
  x = p0.df$trait1[p0.df$ind %in% ctls.sensitiv],
  y = p0.df$trait2[p0.df$ind %in% ctls.sensitiv],
  pch = 17, col = "green"
)
legend("topleft",
  legend = c("resistant", "sensitive"), pch = c(19, 17),
  col = "green", bty = "n"
)
```

## Phenotype controls

Make and save phenotypes of controls in several additionnal plots the last few years to give a good reference to the players:
```{r phenos_controls, eval=TRUE}
nb.years.per.ctl <- 4
nb.plots.per.ctl <- 5
phectls.df <- NULL
latest.year <- as.numeric(levels(p0.df$year)[nlevels(p0.df$year)])
for (j in 1:nb.years.per.ctl) {
  year <- latest.year - nb.years.per.ctl + j
  phectl.df <- makeDfPhenos(
    ind.ids = controls,
    nb.plots.per.ind = rep(
      nb.plots.per.ctl,
      length(controls)
    ),
    year = year,
    pathogen = ifelse((year - first_pheno_year) %% 3 == 0,
      TRUE, FALSE
    )
  )
  phectl.df$plot <- as.factor(as.numeric(phectl.df$plot) +
    nlevels(p0.df$plot))
  phectl <- simulTraits12(
    dat = phectl.df, mu = p0$mu,
    Alpha = p0$Alpha[as.character(year), , drop = FALSE],
    X = coll$genos[levels(phectl.df$ind), thin.snps],
    afs = afs[thin.snps],
    Beta = p0$Beta,
    sigma2 = p0$sigma2, verbose = 1
  )
  phectl$trait3 <- simulTrait3(
    dat = phectl.df,
    X = coll$genos[levels(phectl.df$ind), thin.snps],
    qtn.id = p0$trait3$qtn.id,
    resist.genos = p0$trait3$resist.genos,
    prob.resist.no.qtl = p0$trait3$prob.resist.no.qtl,
    verbose = 0
  )
  phectl.df$trait1.raw <- phectl$Y[, 1]
  phectl.df$trait2 <- phectl$Y[, 2]
  phectl.df$trait3 <- phectl$trait3$y
  phectl.df$trait1 <- phectl.df$trait1.raw
  tmp <- (phectl.df$pathogen & as.logical(phectl.df$trait3))
  if (any(tmp)) {
    phectl.df$trait1[tmp] <- (1 - p0$prop.yield.loss) * phectl.df$trait1[tmp]
  }
  phectl.df <- phectl.df[, -grep("raw", colnames(phectl.df))]
  print(summary(phectl.df))
  if (j == 1) {
    phectls.df <- phectl.df
  } else {
    phectls.df <- rbind(phectls.df, phectl.df)
  }
}
regplot(phectls.df$trait1, phectls.df$trait2,
  xlab = "phectls.df$trait1",
  ylab = "phectls.df$trait2", las = 1
)
fout <- paste0(init.dir, "/Result_phenos_controls.txt.gz")
if (file.exists(fout)) {
  file.remove(fout)
}
```

Enhancement: add spatial coordinates and missing data

Free some memory:
```{r free_mem_phenos, eval=TRUE}
rm(p0)
gc()
```


# Set up database

Create the database:
```{r setup_db}
progress_detail = "Initialise data-base..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)
DATA_DB <- dbname
# db <- dbConnect(SQLite(), dbname = dbname)
# dbDisconnect(db)

db_schema <- read_sql_file("src/db-schema.sql")
db_execute(query = db_schema)

db <- dbConnect(SQLite(), dbname = DATA_DB)
tables <- dbListTables(db)
dbDisconnect(db)
tables
```



## Constants

Fill the table with the game's constants.

```{r table_constants}
constants <- list(
  max.disk.usage = 10,
  version = readLines("VERSION"),
  maxEvalInds = params$maxEvalInds,

  # Haplotypes and genotypes
  nb.chrs = nb.chrs,
  chr.length = L,
  nb.snps = length(thin.snps),

  # phenotypes
  seed.year.effect = 4321,

  # phenotypes trait 1
  mu.trait1 = params.trait1$mu,
  y.min.trait1 = params.trait1$y.min,
  sigma.p2.trait1 = params.trait1$sigma.p2,
  # w.trait1 = params.trait1$w, # this value does not exist
  i.trait1 = params.trait1$i,
  CV.g.trait1 = params.trait1$CV.g,
  sigma.a2.trait1 = params.trait1$sigma.a2,
  h2.trait1 = params.trait1$h2,
  sigma2.trait1 = params.trait1$sigma2,
  sigma.alpha2.trait1 = params.trait1$sigma.alpha2,
  H2.design.trait1 = params.trait1$H2.design,
  sigma.beta2.trait1 = params.trait1$sigma.beta2,

  # phenotypes trait 2
  mu.trait2 = params.trait2$mu,
  y.min.trait2 = params.trait2$y.min,
  sigma.p2.trait2 = params.trait2$sigma.p2,
  CV.g.trait2 = params.trait2$CV.g,
  sigma.a2.trait2 = params.trait2$sigma.a2,
  h2.trait2 = params.trait2$h2,
  sigma2.trait2 = params.trait2$sigma2,
  sigma.alpha2.trait2 = params.trait2$sigma.alpha2,
  H2.design.trait2 = params.trait2$H2.design,
  sigma.beta2.trait2 = params.trait2$sigma.beta2,

  # Initial data
  nb.inds.denovo = nb.inds.denovo,
  nb.snps.hd = nb.snps.hd,
  nb.snps.ld = nb.snps.ld,
  nb.phenotyped.coll = nlevels(p0.df$ind),
  nb.genotyped.coll = length(initially_genotyped_inds),
  max.nb.haplodiplos = params$max.nb.haplodiplos,
  max.nb.pltmatReq = params$max.nb.pltmatReq,
  max.nb.inds = 1000,
  nb.plots = params$nb.plots,
  nb.controls = nb.controls,
  nb.years.per.ctl = nb.years.per.ctl,
  nb.plots.per.ctl = nb.plots.per.ctl,
  register.min.trait1 = register.min.trait1,
  register.min.trait2 = register.min.trait2,

  # Time delays
  first.year = first.year,
  max.upload.pheno.field = "05-31",
  duration.pheno.field = 4, # in months
  duration.pheno.patho = 3,
  duration.allof = 4,
  duration.autof = 4,
  duration.haplodiplo = 12,
  duration.geno.hd = 1,
  duration.geno.ld = 1,
  duration.geno.single = 1,

  # costs
  cost.pheno.field = params$cost.pheno.field,
  cost.pheno.patho = params$cost.pheno.patho,
  cost.allof = params$cost.allof,
  cost.autof = params$cost.autof,
  cost.haplodiplo = params$cost.haplodiplo,
  cost.geno.hd = params$cost.geno.hd,
  cost.geno.ld = params$cost.geno.ld,
  cost.geno.single = params$cost.geno.single,
  cost.register = params$cost.register,
  initialBudget = params$initialBudget * params$cost.pheno.field
)

db_add_constants(constants)
```


## Sessions

Fill with initial game sessions:

```{r table_sessions_fill}
db_add_game_session(NA, "2018-02-15 09:00:00", "2018-02-15 12:00:00", 60, "UTC")
db_add_game_session(NA, "2018-02-15 14:00:00", "2018-02-15 17:00:00", 60, "UTC")
db_add_game_session(NA, "2018-02-16 09:00:00", "2018-02-16 11:40:00", 40, "UTC")
```

## Breeders

Insert the breeder(s):
```{r create_table_breeders}
db_add_breeder("@ALL", NA, NA) # mandatory to add shared initial materials
db_add_breeder("admin", "game master", digest("1234", "md5", serialize = FALSE))
db_add_breeder("test", "tester", digest("", "md5", serialize = FALSE))
```

## Plant material


Save also the collection in the plant material table of each breeder:

```{r fill_table}
initial_pltmat_request_name <- "-- Initial Plant Material --"
db_add_request(id = NA,
               breeder = "@ALL",
               name = initial_pltmat_request_name,
               type = "pltmat",
               game_date = paste0(first.year - nb.years, "-01-01"))

initial_pltmat_request <- data.frame(
  parent1 = ind.ids,
  parent2 = NA,
  child = rownames(coll$genos),
  explanations = "haplodiploidization"
)
initial_plant_mat_request_id <- db_get_game_requests(name = initial_pltmat_request_name)$id
db_add_init_pltmat_req_data(initial_plant_mat_request_id, initial_pltmat_request)
db_add_pltmat(req_id = initial_plant_mat_request_id)


init_individuals <- db_get_individual()
head(init_individuals)
tail(init_individuals)
```

For each individual initially given to the players, save its haplotypes into a file (slow):
```{r save_haplos_in_files}
progress_detail = "Save initial haplotypes..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

system.time({
  haplo_id_association <- mclapply(1:length(ind.ids), function(i) {
    ind.id <- rownames(coll$genos)[i]
    ind <- list(haplos = getHaplosInd(haplos = coll$haplos, ind.name = ind.id))
    f <- file.path(truth.dir, paste0(ind.id, "_haplos.RData"))
    save(ind, file = f)
    return(data.frame(
      id = db_get_individuals_ids(breeder = "@ALL", names = ind.id),
      haplotype_file = f
    ))
    # db_add_ind_haplotype(db_get_individuals_ids(breeder = "@ALL", names = ind.id),
    #                      f)
  }, mc.cores = nb.cores)
  haplo_id_association <- do.call(rbind, haplo_id_association)
  db_add_ind_haplotype(haplo_id_association)
})
```


Save initial phenotypes data.

```{r save_phenotypes_in_db}
additional_column <- which(colnames(p0.df) == "trait1.raw")
initial_phenotypes <- rbind(p0.df[, -additional_column], phectls.df)
db_add_initial_pheno_data(initial_phenotypes)
```

Save initial genotype data in db.

```{r save_genotypes_in_db}
db_add_request(id = NA,
               breeder = "@ALL",
               name = "-- Initial Genotypes --",
               type = "geno",
               game_date = paste0(getBreedingGameConstants()$first.year - 1, "-01-01"))
init_geno_req_id <- db_get_game_requests(name = "-- Initial Genotypes --")$id


init_geno_req <- data.frame(
  ind = initially_genotyped_inds,
  task = "geno",
  details = "hd"
)

db_add_geno_req_data(init_geno_req_id, init_geno_req)

genotype_data_files <- list(hd = initial_genotype_file)
db_add_geno_data(init_geno_req_id, genotype_data_files)


db_mark_as_control(controls)
```






Save also the initial allele frequencies:
```{r save_afs_in_file}
f <- paste0(truth.dir, "/afs0.RData")
afs0 <- afs[thin.snps]
save(afs0, file = f)
```

## Example files

Make an example of a "plant material" file:
```{r ex_cross}
(plant.examples <- makeExamplePlantFile(init.dir))
```

Make an example of an "data" file:
```{r ex_data}
(data.examples <- makeExampleDataFile(init.dir))
```


# Export ZIP archive

Make a zip archive of the whole directory:
```{r make_zip}

progress_detail = "Create zip archive..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)

system.time(
  zip(
    zipfile = paste0(root.dir, ".zip"),
    files = list.files(
      path = root.dir, full.names = TRUE, recursive = TRUE,
      include.dirs = TRUE
    )
  )
)
```

# Specify this new session should be use

```{r, new_session_specification}
progress_detail = "Save session to use..."
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)
data_in_use_file <- file.path(data_root, "data_folder_in_use.txt")
writeLines(session_id, data_in_use_file)
```


# Appendix

```{r info}
t1 <- proc.time()
t1 - t0
print(sessionInfo(), locale = FALSE)

progress_detail = "Game setup script DONE"
progressBar$set(
  value = progressBar$getValue() + 1,
  detail = progress_detail
)
```
